### 2021-06-01 cued/conclusion.tex
Proof-read main report

The Feature-First Block Model (FFBM) introduced 
in this paper is a new generative model for labelled networks,
developed in order to address difficulties of other graphical models 
when testing how vertex features affect community structure. The idea 
is to divide the graph into its most natural partition and test whether 
the vertex features can accurately explain this partition. 
It is relatively easy to find vertex features that are in some way 
correlated with the graphical structure. Nonetheless, only when 
we find the feature that best describes the most pronounced partition 
do we have a stronger case for causation.

Using this new model,
we go on to describe an efficient inference algorithm to sample 
the parameters $\theta$ of the feature-to-block generator. 
This takes the form of a two-level Markov chain,
used to sample the block memberships $b$ and block generator 
parameters $\theta$. These chains can either be executed
in parallel or 
the empirical mean of the $b$-samples can be used
as the input to the $\theta$-chain. The latter option
reduces the variance in our evaluation of the target distribution 
and thus shortens burn-in.

>>>>

The Feature-First Block Model (FFBM) introduced 
in this paper is a new generative model for labelled networks,
developed to address difficulties of other graphical models 
when testing how vertex features affect community structure. The idea 
is to divide the graph into its most natural partition and determine whether 
the vertex features can accurately explain this partition. 
It is relatively easy to find vertex features that are in some way 
correlated with the graphical structure. Nonetheless, only when 
we find the feature that best describes the most pronounced partition 
do we have a stronger case for causation.

Using this new model,
we go on to describe an efficient inference algorithm to sample 
the parameters of the FFBM. 
This takes the form of a two-level Markov chain,
used to sample the block memberships $b$ and block generator 
parameters $\theta$. These chains can either be executed
in parallel or 
the empirical mean of the $b$-samples can be used
as the input to the $\theta$-chain. The latter option
reduces the variance in our evaluation of the target distribution 
and thus shortens burn-in.

### 2021-06-01 cued/experiments.tex
Proof-read main report

Table~\ref{tab:results} summarises the results for each experiment;
a complete list of all the hyperparameters used 
is given in Appendix \ref{appdx:hyperparams}. We also apply the 
dimensionality reduction method 
of Section~\ref{sec:dim-reduction}
to the two higher dimensional datasets (the school and FB egonet). 
For this we use equation~(\ref{eqn:c-star}) with $k=1$,
in order to reduce the dimension from 
$D$ to a desired $D'$. 
We then retrain the feature-block predictor using only the retained 
feature set $\Dcal'$, and report the log-loss over the training and 
test sets for the reduced classifier -- 
denoted $\bar{\Lcal}_0'$ and $\bar{\Lcal}_1'$ respectively. 
These values are also shown in Table~\ref{tab:results}.

@@

However, from Figure~\ref{fig:polbooks-accuracy} we see that block 1 has low accuracy. This suggests that detected block 1 is not solely composed of ``neutral" books but also contains some ``liberal" and ``conservative" authors. Indeed, by examining Figure~\ref{fig:polbooks-graph} we see that block 1 is effectively a buffer between blocks 2 and 3 -- there are very few edges between blocks 2 and 3. It is therefore not surprising that some books from either side leak into block 1. Perhaps, the three distinct categories (``conservative", ``liberal", ``neutral") are too coarse a measure of political affiliation; it may be the case that the nominally ``conservative" books found in block 1 are in fact more centre-right. In the absence of more granular labels, we cannot test this hypothesis. Nevertheless, political affiliation encoded as 3 distinct labels remains a fantastic predictor of network structure.

@@

When the feature dimension is very large, it becomes increasingly likely that 
a particular feature may uniquely identify a small set of nodes. If these nodes 
are all part of the same community, then the classifier may overfit for that 
particular parameter. The regularisation term imposed by the prior goes some 
way towards alleviating this problem. Nevertheless, we see in 
Figure~\ref{fig:fb-null} that the feature \verb*|birthday-5| has a very 
high weight as it relates to block 1 -- but it is highly unlikely
that birthdays determine graphical structure. This issue could have been alleviated by choosing $\Xcal=\{-1, +1\}$ such that having a feature turned off supports excluding a vertex from the block. However, Figure~\ref{fig:fb-null} demonstrates why we made the deliberate choice $\Xcal=\{0,1\}$, as argued in Appendix \ref{appdx:dimension}. Because of the choice $\Xcal=\{0,1\}$, block 1 is allowed to contain 3 distinct feature categories rather than just one. We accept this tradeoff: risking spurious high magnitude components (e.g. \verb*|birthday-5|) for the benefit of allowing a top-level block span multiple feature-groups. Nevertheless, one would have to apply a hierarchical approach to subdivide each top-level block into sub-blocks to test whether the constituent feature-groups determine structure at the lower-level.

>>>>

Table~\ref{tab:results} summarises the results for each experiment;
a complete list of all the hyperparameters used 
is given in Appendix \ref{appdx:hyperparams}. We also apply the 
dimensionality reduction method 
of Section~\ref{sec:dim-reduction}
to the two higher dimensional datasets (the school and FB egonet). 
For this we use equation~(\ref{eqn:c-star}) with $k=1$,
in order to reduce the dimension from 
$D$ to a desired $D'$. 
We then retrain the feature-block predictor using only the retained 
feature set $\Dcal'$, and report the log-loss over the training and 
test sets for the reduced classifier -- 
denoted $\bar{\Lcal}_0'$ and $\bar{\Lcal}_1'$ respectively. 
These values are also given in Table~\ref{tab:results}.

@@

However, from Figure~\ref{fig:polbooks-accuracy} we see that block 1 has low accuracy. 
This suggests that detected block 1 is not solely composed of ``neutral" books but also 
contains some ``liberal" and ``conservative" authors. Indeed, by examining 
Figure~\ref{fig:polbooks-graph} we see that block 1 is effectively a buffer between 
blocks 2 and 3; there are very few edges between blocks 2 and 3. It is therefore not 
surprising that some books from either side leak into block 1.
Perhaps, the three distinct categories (``conservative", ``liberal", ``neutral") are 
too coarse a measure of political affiliation; it may be the case that the nominally 
``conservative" books found in block 1 are in fact more centre-right. 
In the absence of more granular labels, we cannot test this hypothesis. 
Nevertheless, political affiliation encoded as 3 distinct labels 
remains a fantastic predictor of network structure.

@@

When the feature dimension is very large, it becomes increasingly likely that 
a particular feature may uniquely identify a small set of nodes. If these nodes 
are all part of the same community, then the classifier may overfit for that 
particular parameter. The regularisation term imposed by the prior goes some 
way towards alleviating this problem. Nevertheless, we see in 
Figure~\ref{fig:fb-null} that the feature \verb*|birthday-5| has a very 
high weight as it relates to block 1 -- but it is highly unlikely
that birthdays determine graphical structure. This issue could have been alleviated by 
choosing $\Xcal=\{-1, +1\}$ such that having a feature turned off supports excluding 
a vertex from the block. However, Figure~\ref{fig:fb-null} demonstrates why we made 
the deliberate choice $\Xcal=\{0,1\}$, as argued in Appendix \ref{appdx:dimension}. 
Because of the choice $\Xcal=\{0,1\}$, block 1 is allowed to contain 3 distinct feature 
categories rather than just one. We accept this trade-off: risking spurious high magnitude 
components (e.g. \verb*|birthday-5|) for the benefit of allowing a top-level block to span 
multiple feature-groups. Nevertheless, one would have to apply a hierarchical approach 
to subdivide each top-level block into sub-blocks to test whether the constituent 
feature-groups determine structure at the lower-level.

### 2021-06-01 cued/ffbm.tex
More appendix
Proof-read main report

Let $N$ denote the number of nodes, $B$ the number of blocks in the graph,
and ${\cal X}$ the set of values each feature can take.
We define the vector $x_i \in \Xcal^D$ as the feature vector of vertex $i$, 
where $D$ is the number of features associated with each vertex.
For the datasets we analyse, we deal with binary feature flags so $\Xcal = \{0, 1\}$. We write $X$ for the $N\times D$ {\em feature matrix} containing
the feature vectors $\{x_i\}_{i=1}^{N}$ 
as its rows.

For the FFBM, we start with the feature matrix $X$ and generate a random
vector of block memberships $b \in [B]^N$. For each vertex $i$, the
block membership $b_i\in[B]$ is generated based on the feature
vector $x_i$, independently between vertices. The conditional
distribution of $b_i$ given $x_i$ also depends on a collection
of weight vectors $\theta=\{w_k\}_{k=1}^B$, where each
$w_k$ has dimension $D$. Specifically, 
the distribution of $b$ given $X$ and $\theta$ is,
%
\begin{equation}
	p(b| X, \theta) = \prod_{i=1}^{N} p(b_i | x_i, \theta) = \prod_{i=1}^{N} \phi_{b_i} (x_i; \theta)
	= \prod_{i=1}^{N} \frac{\exp\left(w_{b_i}^T x_i\right)}{\sum_{k=1}^{B} \exp \left( w_k^T x_i\right)}.
\end{equation}
%
$\phi_{b_i}$ is the form of a softmax activation function; we deliberately exclude a bias term to ensure that we only leverage feature information. 
We will later also find it convenient
to write the parameters $\theta$ as a $B \times D$ matrix of weights $W$.

@@

To complete the description of our Bayesian framework,
priors on $\theta$ and $\psi$ must also be specified. 
We place a Gaussian prior on $\theta$ with zero mean
and covariance matrix $\sigma^2_\theta I$, so that
each element of $\theta$ has an independent ${\cal N}(0,\sigma_\theta^2)$
prior, with hyperparameter $\sigma_\theta^2$:
%
\begin{equation}
	p(\theta) \sim \Gaussian \left( \theta ; 0, \sigma_\theta^2 I \right).
	\label{eqn:theta-prior}
\end{equation}
%
Interestingly, this choice of prior leads to a particularly
simple distribution on the block membership vector $b$ given $X$.
We show in Appendix~\ref{appdx:b|x} that,
after integrating out $\theta$,
$b$ is uniformly distributed given $X$:
%
\begin{equation}
	p(b | X) = \int p(b | X, \theta) p(\theta) d\theta = B^{-N}.
	\label{eqn:b-pseudo-prior}
\end{equation}
%
This is an important simplification as evaluating $p(b | X)$ does not require an expensive Monte Carlo integration over $\theta$ nor does it require the exact value of $X$. 
\citet{Peixoto-Bayesian-Microcanonical} proposes careful choices for 
the priors on the additional microcanonical SBM parameters $\psi$, which we adopt. 
The idea is to write the joint distribution on $(b, e, k)$ as a product of 
conditionals, $p(b, e, k) = p(b) p(e | b) p(k | e, b)= p(b) p(\psi | b)$. 
In our case, conditioning on $X$ is also necessary, leading to,
%
\begin{equation}
	p(b, \psi | X) = p(b | X) p(\psi | b, X) = p(b | X) p(\psi | b),
	\label{eqn:joint-pseudo-prior}
\end{equation}
%
where we used the fact $\psi$ and $X$ are conditionally 
independent given $b$.  The exact form of the prior
$p(\psi|b)$ is described in Appendix \ref{appdx:prior}.
All that concerns the main argument is that it has
an easily computable form.

>>>>

Let $N$ denote the number of vertices, $B$ the number of blocks in the graph,
and ${\cal X}$ the set of values each feature can take.
We define the vector $x_i \in \Xcal^D$ as the feature vector of vertex $i$, 
where $D$ is the number of features associated with each vertex.
For the datasets we analyse, we deal with binary feature flags so $\Xcal = \{0, 1\}$. We write $X$ for the $N\times D$ {\em feature matrix} containing
the feature vectors $\{x_i\}_{i=1}^{N}$ 
as its rows.

For the FFBM, we start with the feature matrix $X$ and generate a random
vector of block memberships $b \in [B]^N$. For each vertex $i$, the
block membership $b_i\in[B]$ is generated based on the feature
vector $x_i$, independently between vertices. The conditional
distribution of $b_i$ given $x_i$ also depends on a collection
of weight vectors $\theta=\{w_k\}_{k=1}^B$, where each
$w_k$ has dimension $D$. Specifically, 
the distribution of $b$ given $X$ and $\theta$ is,
%
\begin{equation}
	p(b| X, \theta) = \prod_{i \in [N]} p(b_i | x_i, \theta) = \prod_{i \in [N]} \phi_{b_i} (x_i; \theta)
	= \prod_{i \in [N]} \frac{\exp\left(w_{b_i}^T x_i\right)}{\sum_{k \in [B]} \exp \left( w_k^T x_i\right)}.
\end{equation}
%
Note that $\phi_{b_i}$ has the form of a softmax activation function;
we deliberately exclude a bias term to ensure that we only leverage feature information.
This decision is justified in greater detail in Appendix~\ref{appdx:dimension}.
We will later also find it convenient
to write the parameters $\theta$ as a $B \times D$ matrix of weights $W$.

@@

To complete the description of our Bayesian framework,
priors on $\theta$ and $\psi$ must also be specified. 
We place a Gaussian prior on $\theta$ with zero mean
and covariance matrix $\sigma^2_\theta I$, so that
each element of $\theta$ has an independent ${\cal N}(0,\sigma_\theta^2)$
prior, with hyperparameter $\sigma_\theta^2$:
%
\begin{equation}
	p(\theta) \sim \Gaussian \left( \theta ; 0, \sigma_\theta^2 I \right).
	\label{eqn:theta-prior}
\end{equation}
%
Interestingly, this choice of prior leads to a particularly
simple distribution on the block membership vector $b$ given $X$.
We show in Appendix~\ref{appdx:b|x} that,
after integrating out $\theta$,
$b$ is uniformly distributed given $X$:
%
\begin{equation}
	p(b | X) = \int p(b | X, \theta) p(\theta) d\theta = B^{-N}.
	\label{eqn:b-pseudo-prior}
\end{equation}
%
This is an important simplification as evaluating $p(b | X)$ does not require an expensive Monte Carlo integration over $\theta$ nor does it require the exact value of $X$. 
\citet{Peixoto-Bayesian-Microcanonical} proposes careful choices for 
the priors on the additional microcanonical SBM parameters $\psi$, which we adopt. 
The idea is to write the joint distribution on $(b, e, k)$ as a product of 
conditionals, $p(b, e, k) = p(b) p(e | b) p(k | e, b)= p(b) p(\psi | b)$. 
In our case, conditioning on $X$ is also necessary, leading to,
%
\begin{equation}
	p(b, \psi | X) = p(b | X) p(\psi | b, X) = p(b | X) p(\psi | b),
	\label{eqn:joint-pseudo-prior}
\end{equation}
%
where we used the fact $\psi$ and $X$ are conditionally 
independent given $b$.  The exact form of the prior
$p(\psi|b)$ is described in Appendix \ref{appdx:sbm-prior}.
All that concerns the main argument is that it has
an easily computable form.
